---
title: "Lizard Reduce"
output: html_notebook
---
#This file takes a mapped data file, reduces it and export the reduced file for later analysis

```{r setup,include=FALSE}
# rm(list=ls()) #clears the environment
# filefolder="G:/Lizard work for the road/excel files"
filefolder="C:/Users/test/Documents/Chris/Research/Sceloporus/Caudal Autotomy/Demography"
knitr::opts_knit$set(root.dir = normalizePath(filefolder)) 
#this is an option we have to use for RMD notebooks otherwise the wd is reset after each chunk is executed
#setwd(filefolder) #this can be used instead for a regular rscript
# options(pkgType = "source")
mypackages=c("ggplot2","tables","knitr","xlsx","rJava") 
neededpackages=mypackages[which(!is.element(el=mypackages,set=installed.packages()))]
if(length(neededpackages>0)){
    install.packages(neededpackages)
}else{
    warning("All packages already installed")
}
lapply(mypackages,function(x) library(x,character.only = T))
```
Read in a mapped data file and check names
```{r}
mdata<-read.xlsx2("mapped data.xlsx",sheetIndex = 1)
names(mdata)
str(mdata)
```
Create a review data frame and export to an excel file.
```{r}
reviewdata<-data.frame(mdata[which((mdata$review_new.recap==T|mdata$review_painted==T|mdata$review_rtl==T|mdata$review_sex==T|mdata$review_species==T)&(mdata$drop_morphometrics==F & mdata$drop_species==F)),])
str(reviewdata)
unique(reviewdata$species)
write.xlsx2(reviewdata,"reviewdata.xlsx",showNA=T)
```
RESUME HERE. Everything below this point was copied from file_prep_and_analysis_2.Rmd before it became file prep and Mapping

##Assign variables(columns) to appropriate classes, where necessary - this can also wait until data are merged into one df

#Resume here.  we need to review data flagged by filters - Map Phase has ended.  Now we begin the Reduce phase.  In this phase, we'll reduce the size of the dataframe to only include those data which fit our criteria.
```{r, include=FALSE}
#Eliminate entries for species other than Sj and Sv
##Eliminate entries without toes
ind_toes<-which(!is.na(mycombfile[,"toes"])==T)
mycombfile<-mycombfile[ind_toes,]
nrow(mycombfile)


# unique_id<-vector(mode = "character",length = nrow(mycombfile))

# names(mycombfile)[length(mycombfile)]<-"unique_id"
names(mycombfile)[which(names(mycombfile)=="toes")]<-"toes_orig"
mycombfile[,"toes"]<-mycombfile[,"toes_orig"]

my_a<-grep(pattern = "a",x = mycombfile["toes"],ignore.case = T)
my_b<-grep(pattern = "b",x = mycombfile["toes"],ignore.case = T)

mycombfile[my_a,"unique_id"]<-"a"
mycombfile[my_b,"unique_id"]<-"b"
table(mycombfile$unique_id)
#eliminate letters from toes 

#fix this, gsub calls don't seem to work
for(i in 1:nrow(mycombfile)){
        mycombfile[i,"toes"]<-gsub(pattern = "-a|-b| a| b|a|b",replacement = "",x = mycombfile[i,"toes"] ,fixed = F)
}

##Standardizing toes by making separators into " " 
##replace single quotes and parentheses with nothing in toes column

for(i in 1:nrow(mycombfile)){
        mycombfile[i,"toes"]<-gsub(pattern="'|\\(|\\)|\\.000000|\\?|-\\?",replacement= "",x=mycombfile[i,"toes"],fixed = F)
}

## replace single spaces with a dash in toes column

```
 
 Now we fix "j" and "v" replacing them with "sj" and "sv", respectively.
```{r}
for(j in 1:nrow(mycombfile)){
        if(mycombfile[j,"species"]=="j"){
                mycombfile[j,"species"]<-"sj"
        }
        if(mycombfile[j,"species"]=="v"){
                mycombfile[j,"species"]<-"sv"
        }
}

speciesEntries<-unique(mycombfile[,"species"])
print(speciesEntries)

```

The following chunk of code takes a long time (~2minutes to execute).  Figure out why and speed this up.
```{r, include=FALSE}
for(i in 1:nrow(mycombfile)){
        mycombfile[i,"toes"]<-gsub(pattern="\\/| |\\/20| - |-0",replacement="-",x=mycombfile[i,"toes"],fixed = T)
}

print(nchar(mycombfile[1,"toes"]))
print(mycombfile[1,"toes"])

```


```{r}
for(j in 1:nrow(mycombfile)){
        while(any(str_sub(mycombfile[j,"toes"],1L,1L)=="-| ")){
                ##find any dashes that are the first character in an entry
                ind_d1<-which(str_sub(mycombfile[j,"toes"],1L,1L)=="-| ")
                ##removes 1st dash in entries 
                mycombfile[ind_d1,"toes"]<-str_sub(mycombfile[ind_d1,"toes"],2)
        }
        while(any(str_sub(mycombfile[,"toes"],-1L,-1L)=="-| ")){
                ##find any dashes that are the last character in an entry
                ind_d2<-which(str_sub(mycombfile[,"toes"],-1L,-1L)=="-| ")
                ##removes trailing dash in entries 
                mycombfile[ind_d2,"toes"]<-str_sub(mycombfile[,"toes"][ind_d2],1L,-2)
        }
}
```

```{r}
#FIX this part of the code currently
##create column with binary representations of toe loss status for all 20 toes
## 1= toe loss; 0 = no loss; e.g. is only toe 20 is missing, the 20th character 
## will be 1 and all other chracters will be 0
##create new column with class character containing a string of twenty zeros in each entry

# rm(ind_noNA)
# rm(ind_toes)
# rm(ind_SjSv)

mycombfile[,"toes_bin"]<-c()
mycombfile[,"toes_bin"]="00000000000000000000"

test<-list()
for(j in nrow(mycombfile)){
        test[[j]]<-strsplit(mycombfile[j,"toes"], split = "-")
        }
test[1]

for(j in 1:nrow(mycombfile)){
        my_toes_bar<-unlist(strsplit(as.character(mycombfile["toes"]), split = " |-"))#split arguement 
        ntoe<-as.integer(my_toes_bar[])
        substr(mycombfile[j,"toes_bin"],ntoe,ntoe)<-"1"
        }
        # rm(my_toes_bar)
        # rm(ntoe)
}
length(mycombfile[which(is.na(mycombfile$toes_bin)),"toes_orig"])
head(mycombfile[which(is.na(mycombfile[,"toes_bin"])),c("toes","toes_bin")])
```

```{r}
#lapply(mydata_vars,function(x) table(x[,"toes_bin"]))# This is a check to see if the binary toe clip representations were generated properly for all files. There should be a lot of these and mostly unique within species.  This should, in fact, match the result of a similar operation on the numerical toes.
```

```{r}
#deal with sex
table(mycombfile$sex)

## convert definite f values to f and m values to m
mycombfile$sex<-gsub(pattern = "f ",replacement = "f",x = mycombfile$sex,ignore.case = T)
mycombfile$sex<-gsub(pattern = "m |male|n",replacement = "m",x = mycombfile$sex,ignore.case = T)

table(mycombfile$sex)
```

The operations here need to be determined based on the results of manual check(s).
```{r}
## check
table(mycombfile$sex)

##create a second sex column with only definite values
mycombfile$sex_certain<-mycombfile$sex
mycombfile$sex_certain<-gsub(pattern = "\\?|f\\?",replacement = NA,x = mycombfile$sex_certain,ignore.case = T)

## check
table(mycombfile$sex_certain)
```

We follow a similar proces for cleaning species data.
```{r}
# deal with species
table(mycombfile$species)

mycombfile[which(mycombfile$species=="sj?"),]
mycombfile[which(mycombfile$toes=="4-10-19"),] 
####decision: convert "sj?" to "sv"
### convert definite sj values to sj
mycombfile$species<-gsub(pattern = "sj\\?|sv ",replacement = "sv",x = mycombfile$species,ignore.case = T)
mycombfile$species<-gsub(pattern = "sj ",replacement = "sj",x = mycombfile$species,ignore.case = T)
table(mycombfile$species)#check
```

##Handling Duplicate Data
Here we identify instances where the duplicate combinations of the critical variables species, sex, and toes occur within the same year. First we will create a categorical variable, tailcondition, which is determined by the magnitude and sign of the value in the RTL column. 
```{r}
mycombfile$tailcondition<-"Intact"
mycombfile[which(mycombfile$rtl>0),"tailcondition"]<-"Regrown"
mycombfile[which(mycombfile$rtl<0),"tailcondition"]<-"Autotomized"

# mycomparables<-mycombfile[,c("species","toes_orig","sex_certain","year")]


```
Then we will create a variable that identifies whether or not an entries values for the critial variabels are found elsewhere within the entries for that year.
```{r}
mydupvalues<-mycombfile[which(duplicated(mycomparables)==T|duplicated(mycomparables,fromLast = T)==T),]#returns records for duplicated values
nrow(mydupvalues)#Check

mydupvalues<-mydupvalues[with(mydupvalues,order(year,sex_certain,toes_orig)),]#orders mydupvalues by year sex and toes
mycombfile$dupInYear<-"No" #Create a new variable identifying duplicate data
mycombfile[which(duplicated(mycomparables)==T|duplicated(mycomparables,fromLast = T)==T),"dupInYear"]<-"Yes"
```

write file to an .xlsx
```{r}
filename<- paste("Combined",min(mycombfile$year),max(mycombfile$year),".xlsx",sep = "_")
write.xlsx(x = mycombfile,file =filename)
```

