---
title: "Summary Analysis of Lizard Data"
output:
  html_document:
    df_print: paged
---

The input to this script is a cleaned dataset

In this chunk we set the path, library packages and read in data
```{r,setup}

wd = 'S:/Chris/tailDemography/data'
packages = c('ggplot2', 'knitr')
for (package in packages){
  if(package %in% installed.packages()){
    print(paste0(package,' is already installed.'))
    }else{
      install.packages(package)
    }
}
knitr::opts_knit$set(root.dir = normalizePath(wd))
if(getwd()==wd){
    sprintf('You are already in the desired working directory:
            %s.',getwd())
}else{
    setwd(wd)
    sprintf('Working directory is now %s',wd)
}
```

```{r}
if('ggplot2' %in% installed.packages()){
    print('ggplot2 is already installed.')
    }else{
        install.packages('ggplot2')
    }
```

```{r}
library(ggplot2)
```


```{r}
# df = read.csv('../data/Cleaned Sj data.csv')
df = read.csv('./cleaned CC data 2000-2017.csv')
head(df)

```
Filter data for analyses
```{r}
#remove non sj and sv species and non m and f sexes
df = df[which((df$species == 'j') & (df$sex == 'm'|df$sex =='f')),]
#create data set that onyl includes first sightings of animals
df_first = df[which(df$capture==1),]
df_recap = df[which(df$capture>1),]
df_first
df_recap
unique(df$capture)
```

# Looking at first captures in the  data set
## Plot SVL vs TL by year and Autotomy status
- We see more individuals are intact at first capture than are autotomized
```{r}
svlTlall = ggplot(aes(x = svl, y = tl), data = df_first)
fig_svlTLall = svlTlall + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle('TL vs SVL by Sex and Autotomy Status') +ylab('TL in mm') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth(aes(), data = df_first[which(df_first$sex=='f'),],method = lm,colour="red")  + geom_smooth(aes(), data = df_first[which(df_first$sex=='m'),],method = lm,colour="blue") + theme(plot.title = element_text(hjust = 0.5))
fig_svlTLYear= fig_svlTLall +facet_wrap(~year)
ggsave('../figures/svlTL_/SVLvTLall.png',fig_svlTLall)
# ggsave('../figures/svlTL_/SVLvTLYear.png', fig_svlTLYear)
print(fig_svlTLall)
# print(fig_svlTLYear)
```

```{r}
for(year in unique(df_first$year)){
    svlTl = ggplot(aes(x = svl, y = tl), data = df_first[which(df_first$year == year),])
fig_svlTl = svlTl + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle(paste('TL vs SVL by Sex and Autotomy Status in',year,sep = ' ')) +ylab('TL in mm') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth(aes(), data = df_first[which(df_first$sex=='f'),],method = lm,colour="red")  + geom_smooth(aes(), data = df_first[which(df_first$sex=='m'),],method = lm,colour="blue") + theme(plot.title = element_text(hjust = 0.5))
filename = paste('svlTl_',year,'.png',sep = '')
mypath = paste('../figures/svlTL_',filename,sep = '')
print(mypath)
ggsave(mypath,fig_svlTl)
print(fig_svlTl)
}


```

## Plot Recaptured Lizards by sex and year
- One question is why there are any individuals here in blue
```{r}
## need to show y labels
recapSexYr = ggplot(aes(x = factor(year),  fill= newRecap), data = df_first[which(df_first$newRecap=='new'|df_first$newRecap=='recap'),])
fig_recapSexYr = recapSexYr + geom_bar(aes(fill = newRecap), position = 'fill') + theme(axis.text.x = element_text(angle = 45)) + xlab("Year") + ylab("Proportion")+facet_wrap(~sex,nrow = 2)
ggsave('../figures/recapSex_/recapSexYr.png',fig_recapSexYr)
print(fig_recapSexYr)
```
## plot svl by mass and year
```{r}
svlMassYearall = ggplot(aes(x = svl, y = mass), data = df_first)
fig_svlMassYearall = svlMassYearall + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle('Mass vs SVL by Sex and Autotomy Status') +ylab('Mass in g') + xlab('SVL in mm')+ geom_vline(xintercept = 54)   + geom_smooth(aes(), data = df_first[which(df_first$sex=='f'),],colour="red")  + geom_smooth(aes(), data = df_first[which(df_first$sex=='m'),],colour="blue")+ theme(plot.title = element_text(hjust = 0.5))
ggsave('../figures/svlMassYearall.png',fig_svlMassYearall)
print(fig_svlMassYearall)
```

```{r}
fig_svlMassYear= fig_svlMassYearall +facet_wrap(~year)
ggsave('../figures/svlMassYear.png',fig_svlMassYear)
print(fig_svlMassYear)
```

```{r}
for(year in unique(df_first$year)){
    svlMass = ggplot(aes(x = svl, y = mass), data = df_first[which(df_first$year == year),])
fig_svlMass = svlMass + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle(paste('Mass vs SVL by Sex and Autotomy Status',year,sep = '')) +ylab('Mass in g') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth(aes(), data = df_first[which(df_first$sex=='f'),],colour="red")  + geom_smooth(aes(), data = df_first[which(df_first$sex=='m'),],colour="blue")+ theme(plot.title = element_text(hjust = 0.5)) + xlim(c(0,max(df_first$svl)+10))
filename = paste('svlMass_',year,'.png',sep = '')
mypath = paste('../figures/svlMass_/',filename,sep = '')
print(mypath)
ggsave(mypath,fig_svlMass)
print(fig_svlMass)
}
```

# Looking at recaps in the data set

## Plot SVL vs TL by year and Autotomy status
- We see more individuals are intact at first capture than are autotomized
```{r}
svlTlall = ggplot(aes(x = svl, y = tl), data = df_recap)
fig_svlTLall = svlTlall + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle('TL vs SVL by Sex and Autotomy Status') +ylab('TL in mm') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth(method = lm) + theme(plot.title = element_text(hjust = 0.5))
fig_svlTLYear= fig_svlTLall +facet_wrap(~year)
ggsave('../figures/svlTL_/SVLvTLall_recap.png',fig_svlTLall)
ggsave('../figures/svlTL_/SVLvTLYear_recap.png', fig_svlTLYear)
print(fig_svlTLall)
print(fig_svlTLYear)
```

```{r}
for(year in unique(df_recap$year)){
    svlTl = ggplot(aes(x = svl, y = tl), data = df_recap[which(df_recap$year == year),])
fig_svlTl = svlTl + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle(paste('TL vs SVL by Sex and Autotomy Status in',year,sep = ' ')) +ylab('TL in mm') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth(method = lm) + theme(plot.title = element_text(hjust = 0.5))
filename = paste('svlTl_',year,'_recap.png',sep = '')
mypath = paste('../figures/svlTL_',filename,sep = '')
print(mypath)
ggsave(mypath,fig_svlTl)
print(fig_svlTl)
}


```

## Plot Recaptured Lizards by sex and year
- One question is why there are any individuals here in blue
```{r}
## need to show y labels
recapSexYr = ggplot(aes(x = factor(year),  fill= newRecap), data = df_recap[which(df_recap$newRecap=='new'|df_recap$newRecap=='recap'),])
fig_recapSexYr = recapSexYr + geom_bar(aes(fill = newRecap), position = 'fill') + theme(axis.text.x = element_text(angle = 45)) + xlab("Year") + ylab("Proportion")+facet_wrap(~sex,nrow = 2)
ggsave('../figures/recapSex_/recapSexYr_recap.png',fig_recapSexYr)
print(fig_recapSexYr)
```
## plot svl by mass and year
```{r}
svlMassYearall = ggplot(aes(x = svl, y = mass), data = df_recap)
fig_svlMassYearall = svlMassYearall + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle('Mass vs SVL by Sex and Autotomy Status') +ylab('Mass in g') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth() + theme(plot.title = element_text(hjust = 0.5))
fig_svlMassYear= fig_svlMassYearall +facet_wrap(~year)
ggsave('../figures/svlMassYearall_recap.png',fig_svlMassYearall)
ggsave('../figures/svlMassYear_recap.png',fig_svlMassYear)
print(fig_svlMassYearall)
print(fig_svlMassYear)

for(year in unique(df_recap$year)){
    svlMass = ggplot(aes(x = svl, y = mass), data = df_recap[which(df_recap$year == year),])
fig_svlMass = svlMass + geom_point(aes(color = factor(sex), shape = factor(autotomized))) + scale_shape( solid = FALSE) + ggtitle(paste('Mass vs SVL by Sex and Autotomy Status',year,sep = '')) +ylab('Mass in g') + xlab('SVL in mm')+ geom_vline(xintercept = 54)  + geom_smooth() + theme(plot.title = element_text(hjust = 0.5)) + xlim(c(0,max(df_recap$svl)+10))
filename = paste('svlMass_',year,'_recap.png',sep = '')
mypath = paste('../figures/svlMass_/',filename,sep = '')
print(mypath)
ggsave(mypath,fig_svlMass)
print(fig_svlMass)
}